---
title: "Course-Project"
output: html_document
---

## Introduction
Here I will analyze the relationship between transmission type (automatic vs
manual) on the fuel efficiency (miles per gallon) of the cars in the mtcars data
set. First I will perform an exploratory data analysis, then regression modeling
and statistical inference, check the reliability of the model with residual
diagnostics, and finally present conclusions and assessment of uncertainty.

## Exploratory Analysis
First, to get a rough idea of what kind of data and how much there is in the
data set, look at the head of mtcars

```{r}
head(mtcars)
```

Looks like the transmission type is recorded in the field "am", as a binary
value where 1 means manual, 0 means automatic. This data is already tidy, so for
a preliminary analysis we can just use a simple linear model of the correlation
we hope to test.

```{r}
mdl_a <- lm(mpg~am, mtcars)
summary(mdl_a)$coef
```

This suggests that naively, manual transmissions seem to get 7 more miles per
gallon. However, there may be bias, since other variables in this data seem 
likely to be correlated. To check for the possibility of bias, examine the
correlation matrix, using absolute value to check strength of correlations:

```{r}
heatmap(abs(cor(mtcars)))
```

From this, it looks like mpg and am are both significantly correlated with wt,
cyl, and disp, creating the possibility of bias if these variables are omitted.
am is also significantly correlated with drat and gear, but this second group
is less correlated with mpg, making it unlikely to cause bias.

## Regression Modeling and Statistical Inference
First, create a linear model to regress for the relationship between mpg and am,
with wt, cyl, and disp included as confounding variables.

```{r}
mdl_awcd <- lm(mpg~am+wt+cyl+disp, mtcars)
summary(mdl_awcd)$coef
```

The coefficient for disp seems quite low, suggesting it may be unnecessary if it
is sufficiently dependent on the other two confounders to not introduce its own
effect. To test this, use ANOVA with various combinations of confounders.

```{r}
mdl_aw <- lm(mpg~am+wt, mtcars)
mdl_awc <- lm(mpg~am+wt+cyl, mtcars)
anova(mdl_a, mdl_aw, mdl_awc, mdl_awcd)
```

This result confirms that the am and wt confounders are necessary, but disp is
not. Thus, to minimize variance inflation, use only wt and cyl as confounders
for further analysis and visualization (mdl_awc).

## Residual Diagnostics
Plot the residuals of the model for any signs that suggest an incorrect model.

```{r}
res <- resid(mdl_awc)
plot(res)
```

Since there is no sign of heteroscedasticity or a visible trend, the model is
likely acceptable.

## Conclusions and Uncertainty
In the exploratory analysis, it initially seemed that transmission type had a 
major effect on mpg, since the coefficient was `r summary(mdl_a)$coef[2,1]` with
a p-value of `r summary(mdl_a)$coef[2,4]`. 

However, after checking for bias and removing the confounding effects of wt and
cyl, the coefficient became `r summary(mdl_awc)$coef[2,1]`, with a p-value of
`r summary(mdl_awc)$coef[2,4]`. This is neither strong nor significant. Thus, 
there is no evidence that transmission type affects gas mileage, even though it 
may appear so before removing bias.