---
title: "Pitch"
output: slidy_presentation
---

## Overview
My project is a simple web app that does automatic encoding and decoding of 
a few simple ciphers, namely the Caesar, Atbash, and Vignere Ciphers. Each tab
has the appropriate input fields for its corresponding cipher, a pair of radio
buttons to control whether it encodes or decodes, and simply outputs the result
to the screen, where the user can read or copy it freely.

## The Caesar Cipher
This is a classic that's a fun puzzle for kids just getting into cryptography.
The code is simple too - all it does is sort the message into uppercase and 
lowercase, translate into integers and move them down to the 0-25 range, apply
the key, apply a modulo to make sure everything stays within the alphabet, and 
move back up the appropriate ranges before converting back to characters for
output.

```{r eval=F}
nums <- utf8ToInt(input$caesarIn)
uppr <- which(nums>=65&nums<=90)
lowr <- which(nums>=97&nums<=122)
key <- input$caesarKey
if (input$caesarED=='Decode') {key <- -key}
nums[uppr] <- nums[uppr] - 65
nums[lowr] <- nums[lowr] - 97
nums[c(uppr, lowr)] <- nums[c(uppr, lowr)] + key
nums[c(uppr, lowr)] <- nums[c(uppr, lowr)]%%26
nums[uppr] <- nums[uppr] + 65
nums[lowr] <- nums[lowr] +97
return(intToUtf8(nums))
```

## The Atbash Cipher
The Atbash Cipher is another classic, and I think actually older than the
Caesar, but a bit less well known. The code is even simpler than the Caesar,
it just sorts into uppercase and lowercase and flips across the appropriate
axes. 

```{r eval=F}
nums <- utf8ToInt(input$atbashIn)
uppr <- which(nums>=65&nums<=90)
lowr <- which(nums>=97&nums<=122)
nums[uppr] <- nums[uppr] + 2*(77.5-nums[uppr])
nums[lowr] <- nums[lowr] + 2*(109.5-nums[lowr])
return(intToUtf8(nums))
```
In fact, it's short enough that I can run a sample in the slide!
```{r}
nums <- utf8ToInt('Look, a demo! With caps and punctuation and stuff! In the slide!')
uppr <- which(nums>=65&nums<=90)
lowr <- which(nums>=97&nums<=122)
nums[uppr] <- nums[uppr] + 2*(77.5-nums[uppr])
nums[lowr] <- nums[lowr] + 2*(109.5-nums[lowr])
print(intToUtf8(nums))
```

## The Vignere Cipher
Finally, the Vignere Cipher adds some sophistication after the simple Caesar and
Atbash, both of which are straightforwardly solvable with frequency analysis or
just brute force guessing. The Vignere was actually thought to be unbreakable
for a whole, until it was... uh... broken. The code is remarkably simple, since
R's method of adding vectors with different lengths is exactly how the Vignere
applies its key! That similarity is actually what gave me the idea for this, and
I actually started with just the Vignere before adding the Caesar and Atbash.
Here's what the code looks like:
```{r eval=F}
nums <- utf8ToInt(input$vignereIn)
uppr <- which(nums>=65&nums<=90)
lowr <- which(nums>=97&nums<=122)
key <- utf8ToInt(tolower(input$vignereKey))-97
if (input$vignereED=='Decode') {key <- -key}
nums[uppr] <- nums[uppr] - 65
nums[lowr] <- nums[lowr] - 97
locs <- c(uppr, lowr)[order(c(uppr, lowr))]
nums[locs] <- nums[locs] + key
nums[locs] <- nums[locs]%%26
nums[uppr] <- nums[uppr] + 65
nums[lowr] <- nums[lowr] +97
return(intToUtf8(nums))
```

## Other Ciphers
There's a bunch of other stuff I could have done, but I didn't really want to do
that much more. The Caesar and Atbash are both specific versions of a more 
general method called the Affine Cipher. The Affine takes two integer keys and
shifts by the first, then flips according to the second. Interestingly, while a
vectorization of the Caesar is called a Vignere, I'm not sure if there's a name
for the vectorized Atbash or Affine. Maybe they're just not as name-worthy
because they weren't as famous historically.

In any case, thank you for reading all the way to the end of my pitch thing!
Have a wonderful day, friend on the other side of the internet.