---
title: "Course-Project-1"
output: html_document
---

## Part 1: Simulation Exercise
### Overview
This section contains simulations of the exponential distribution, measurement
and comparison of sample parameters to theoretical expectations, and a 
demonstration that the distribution of these sample parameters is roughly
Gaussian using both visual and statistical methods.

### Simulations
As instructed in the assignment, simulate the exponential distribution using R's
rexp() function, using samples of 40 exponentials, and simulating 1000 samples.

```{r}
data <- apply(matrix(1:1000), 1, function(x) rexp(40, 0.2))
```

This creates a 40x1000 matrix, where each of the 1000 columns contains a sample
of 40 observations taken from an exponential distribution with lambda=0.2. Next,
take the mean of each sample of 40 by applying the mean function column-wise.

```{r}
meansDist <- apply(data, 2, mean)
```

### Sample Mean vs Theoretical Mean
Each of these 40-observation samples has its own sample mean. To get an idea of
what that sample mean tends to be, find the mean of the sample means:

```{r}
mean(meansDist)
```

In theory, the sample mean should be centered around 1/lambda = 1/0.2 = 5. Thus,
the mean observed here closely matches our theoretical expectations. Some small
difference is to be expected, but this difference will approach zero as the
number of observations per sample or the number of samples approach infinity.

### Sample Variance vs Theoretical Variance
Each sample also has a sample standard deviation. Calculate the mean sample
standard deviation:

```{r}
var(meansDist)
```

This distribution should have a variance close to s^2/sqrt(n). Since the sample
variance is expected to be 1/lambda = 5, and each sample has n=40, the variance
of meansDist should be about 5^2/40 = 25/40 = 0.625. Thus, the result observed
here matches theoretical expectations. Some difference is expected, but this
will go to zero as n goes to infinity.

### Distribution
Visualize the distribution of sample means and check if it is roughly normal:

```{r}
par(mfrow=c(1,2))
hist(apply(data, 2, mean), breaks=20, xlab="Value", ylab="Frequency", main="Histogram of Sample Means")
hist(data[1,], breaks=20, xlab="Value", ylab="Frequency", main="Histogram of Observations")
```

While the histogram of original observations on the right is exponential, the
Central Limit Theorem means that the distribution of sample means on the left is
roughly normal. Besides this visual test, further check normality using the 
Shapiro-Wilk Normality Test, with the R function shapiro.test().

```{r}
shapiro.test(data[1,])
```

This returns an extremely low p-value: thus, the distribution is almost
certainly normal.