---
title: "Course-Project-2"
output: html_document
---

## Part 2: Basic Inferential Data Analysis
### Overview
This section contains an exploratory analysis of the ToothGrowth data in R's
datasets package, a basic summary of the data, and an analysis of the effects of
the supp and dose fields.

### Loading and Exploring the Data
Make sure the datasets package is loaded, and then grab the ToothGrowth data:

```{r}
library(datasets)
tg <- ToothGrowth
```

Next, explore some basic aspects of the data, to get a sense of what's going on.

```{r}
dim(tg) #dimensions of data
head(tg) #examine the first bit of data
apply(tg, 2, class) #classes of the fields
```

### Basic Summary
After very basic exploratory analysis, check the contents of the data itself
```{r}
summary(as.numeric(tg$len))
table(tg$supp)
table(tg$dose)
boxplot(len ~ dose*supp, data=tg)
```

Since supp and dose are discrete while len is continuous, it seems likely that
this is data from an experiment testing different doses of two different 
supplements and measuring tooth length. Visually, dose seems strongly correlated
with greater len. The relationship of sup with len is harder to discern.

### Inferential Statistical Analysis
Use the statistical tools learned in class (t-test, p-values) to discern the 
relationships of supp and dose with len.

```{r}
t.test(tg$len~tg$supp)[c("p.value", "estimate")] #test difference between supp="VC" and supp="OJ"
t.test(subset(tg$len, tg$dose==1), subset(tg$len, tg$dose==0.5), alternative="greater")[c("p.value", "estimate")] #test difference between dose 0.5 and 1
t.test(subset(tg$len, tg$dose==2), subset(tg$len, tg$dose==1), alternative="greater")[c("p.value", "estimate")] #test difference between dose 1 and 2
```

### Conclusions and Underlying Assumptions
The results of the t-tests above are as follows: at an alpha=0.05 level, there
is not sufficient evidence for a difference in population means between the
OJ and VC supplements. However, there is sufficient evidence for differences
between all three dose levels, pairwise. 

These results rely on the following assumptions: observations are iid, groups
are unpaired, variables are roughly normally distributed. There are other
assumptions of course, but these three seem most pertinent.