---
title: "Project-2"
output: html_document
---

## Introduction
Here I will investigate the relationship between city size and GDP, as part of 
an attempt to further explore the claim made in *Growth* by Vaclav Smil, that 
cities have "a disproportionate influence on creating culture, setting 
commercial priorities, influencing political development, supplying an educated
labor force, determining tastes and fashions, and fostering innovation"
(Smil, 332). Much of this is difficult to investigate quantitatively, but GDP
can serve as an extremely rough proxy for these variables. The exact hypothesis
acting as a proxy for Smil's claim is "GDP scales nonlinearly with city size,
with significant positive curvature".
```{r setup, message=F}
library(caret)
library(ggmap)
library(leaflet)
library(magrittr)
library(mice)
library(rvest)
```

## Getting and Pre-Processing Data
First, get the data from Wikipedia. Read it in as an html tree, then get the
appropriate node using the path obtained via Inspect Element on the original
page, then convert to a table. Do this for Wikipedia's lists of cities with
highest GDPs and populations.
```{r get}
gdp_wiki <- 'https://en.wikipedia.org/wiki/List_of_cities_by_GDP' %>%
  url %>%
  read_html() %>%
  html_node('body #content #bodyContent #mw-content-text .mw-parser-output .wikitable') %>%
  html_table(fill=T)
pop_wiki <- 'https://en.wikipedia.org/wiki/List_of_largest_cities' %>%
  url %>%
  read_html %>%
  html_node('body #content #bodyContent #mw-content-text .mw-parser-output .sortable') %>%
  html_table(fill=T)
```
Next, organize the table. Select only relevant columns from each table, renaming
them with tidier names. Merge them by their first columns, which identify each
row. This will return only rows for cities with both population and GDP data.
```{r tidy}
gdp_wiki <- gdp_wiki[, c(1,4,5,6,7,8)]
names(gdp_wiki) <- c("city", "official.nom.gdp", "brook.ppp.gdp", "pwc.ppp.gdp", "mckin.nom.gdp", "other.gdp")
pop_wiki <- pop_wiki[, c(1,4,6,8,10)]
names(pop_wiki) <- c("City", "un_est.pop", "city_proper.pop", "metro_area.pop", "urban_area.pop")
wiki <- merge(gdp_wiki, pop_wiki, by=c(1,1), )
```
Now that the table itself is organized, clean up the data. All the data is in
character format, and most have odds and ends that must be removed before they
can be converted to numeric. Most of these are either commas, citations in [],
or dates in (). After removing these, convert to numeric. Finally, remove
columns which are particularly sparse, and impute missing data using mice.
```{r clean, warning=F, message=F}
wiki <- data.frame(apply(wiki, 2, function(x) gsub(",|\\[.*\\]|\\(.*\\)","",x)))
wiki[,-1] <- apply(wiki[,-1], 2, as.numeric)
apply(wiki, 2, function(x) mean(is.na(x)))
wiki <- wiki[, -c(6, 9)]
wiki <- complete(mice(wiki, print=F))
```
Finally, summarize the data, creating two new fields to hold averages of the
different measures of population and GDP. This makes simple visualization more
straightforward when less detail is needed, and may even provide a more accurate
estimator of the variables these measures are meant to capture.
```{r summarize}
wiki$gdp <- apply(wiki, 1, function(x) mean(as.numeric(x[c(2,3,4,5)])))
wiki$pop <- apply(wiki, 1, function(x) mean(as.numeric(x[c(6,7,8)])))
```

## Exploratory Analysis
Start by checking the correlation matrix - expecting to see two blocks, as GDP
measures and population measures should correlate highly. Next, visualize
distributions of city GDP measures and population measures.
```{r vis}
heatmap(cor(wiki[,-1]))
par(mfrow=c(1,2))
boxplot(wiki[,c(2,3,4,5,9)], main="GDP")
boxplot(wiki[,c(6,7,8,10)], main="POP")
```
Since the distributions of population measures and gdp measures are all roughly
the same shape, the "gdp" and "pop" summary fields will likely be representative
and usable.

## Leaflet Map
Map out the data: here circle size corresponds to the UN's estimate of 
population size, and circle color corresponds to the log of official nominal
GDP.
```{r map, cache=T, message=F}
colorMap <- colorNumeric(c('red', 'blue'), domain=NULL)
geocode(wiki$city) %>%
  leaflet %>%
  addTiles %>%
  addCircles(radius=wiki$pop/30, color=colorMap(log10(wiki$gdp)))
```

## Analysis
Try fitting two regression models: linear and exponential.
```{r model, cache=T}
line <- lm(gdp~pop, data=wiki)
expo <- lm(gdp~log10(pop), data=wiki)
```
To evaluate these models, first visually compare to the plot. Next, examine the
model summaries to determine which is the best mechanistic fit, by Occam's Razor
basically.
```{r eval}
plot(gdp~pop, data=wiki)
abline(line, col='red')
lines(seq(5000000, 30000000, 1000000), predict(expo, data.frame(pop=seq(5000000, 30000000, 1000000))), col='blue')
summary(line)
summary(expo)
```

## Conclusions
Based on this analysis, frankly very little of interest has been found. 
Regarding the original hypothesis, no convincing evidence was found to back up
Smil's claim. However, this is a relatively low-power experiment due to the
multiple layers of proxies and small sample size, so this is not too meaningful.
Interestingly, the curve fit model found a negative curvature, which would imply
diminishing returns on GDP with increasing city population, precisely the
opposite of Smil's claim. No interesting geographic patterns were found either.


All in all, the main conclusion is that city productivity is too complex to be
analyzed with just some Wikipedia tables.